# Statistical checks

Statistical checks involve properties statistical characteristics such 
as the means of columns. These characteristics can be checked for whole
columns or grouped by one or more categorical variables. It is also
possible to use group-wise computed statistics, in validation rules.
For example by taking the ratio between a column and it's group-wise
means, and comparing that with a bound. 

For long-form data it is possible to compare aggregate values with underlying
details. For example to test whether quarterly time series add up to annual
totals. It is also possible to check properties of groups, for example whether
in every household (a group of persons) there is exactly one head of household.

#### Data and packages {-}

In this Chapter we will use the `SBS2000` dataset that comes with `validate`.

```{r}
library(validate)
data(SBS2000)
head(SBS2000, 3)
```

We shall also use the `samplonomy` dataset that also comes with `validate`. See also
\@ref(long-data).

```{r}
data(samplonomy)
head(samplonomy, 3)
```


## Statistical and groupwise characteristics

Any R statement that ultimately is an inequality or inequality check
is a valid validation rule. This means that any statistical calculation
can be input to a rule.

Here we check the mean profit and correlation coefficient between profit and
turnover.
```{r}
rule <- validator(
    mean(profit, na.rm=TRUIE) >= 1
  , cor(turnover, staff, use="pairwise.complete.obs") > 0
)
out <- confront(SBS2000, rule)
# suppress some columns for brevity
summary(out)[1:7]
```

There are a few helper functions to compute group-wise statistics, and to make
comparing values with group aggregates possible.

For example, here we check whether each turnover is less than ten times
the group-wise median.
```{r}
rule <- validator(
  turnover <= 10*do_by(turnover, by=size, fun=median, na.rm=TRUE)
)
out <- confront(SBS2000, rule)
# suppress some columns for brevity
summary(out)[1:7]
```
Here, in the right-hand side of the rule the group-wise median of turnover is
computed. The function `do_by` is very similar to functions such as `tapply`
in base R. The difference is that `do_by` works on vectors only (not on data
frames) and always repeats the values of `fun` so that the length of the output is
equal to the length of the input.
```{r}
medians <- with(SBS2000, do_by(turnover, by=size, fun=median, na.rm=TRUE))
head(data.frame(size = SBS2000$size, median=medians))
```
There are also some convenience functions, including `sum_by`, `mean_by`, `min_by`, and `max_by`.

## Group properties

In this section, we group data by one or more categorical variables and
check for each group whether a rule is satisfied. In particular we are
going to check whether each household in a small dataset has a unique 
'head of household'.

We first create some data with household id (`hhid`) a person id (`person`) and
that person's role in the household (`hhrole`).
```{r}
d <- data.frame(
   hhid   = c(1,  1,  2,  1,  2,  2,  3 )
 , person = c(1,  2,  3,  4,  5,  6,  7 )
 , hhrole = c("h","h","m","m","h","m","m")
)
d
```
With `exists_one()` we can check that there is exactly one person
with the role `"h"` (head) in each household, by grouping on household id.
```{r}
rule <- validator(exists_one(hhrole == "h", by=hhid))
out <- confront(d, rule)
# suppress some columns for brevity
summary(out)
```
We can inspect the results by selecting the violating record groups.
```{r}
violating(d, out)
```
We see that household 1 has two heads of household, while household 3 has no head 
of household. 

To test whether _at least one_ head of household exists, one can use
`exists_any`:

```{r}
violating(d, validator(exists_any(hhrole=="h",by=hhid) ))
```


In the following example we check whether there is exactly one region called Samplonia
for each period and each measure in the `samplonomy` dataset.

```{r}
rule <- validator(exists_one(region=="Samplonia", by=list(period, measure)))
```
The first argument of `exists_one()` is a rule that has to be checked in every group
indicated by the `by` argument. The output is a logical vector with an element for
each record in the dataset under scrutiny. If a group of data fails the test, each record
in that group is indicated as wrong (`FALSE`).

```{r}
out <- confront(samplonomy, rule)
# suppress some columns for brevity
summary(out)
```
Here, there are no groups that violate this assumption. 
```{r}
violating(samplonomy, out)
```




## Aggregates in long-format data

Checking aggregations in long-form format is more involved than for
data in wide format (as in Section \@ref(balance-equalities-and-inequalities)).


Here, we check in the `samplonomy` dataset that for each measure and each
period, the subregional data adds up to the regional data.
```{r}
rules <- validator(
  part_whole_relation(value
    , labels=region
    , whole="Samplonia"
    , part =c("Agria","Induston")
    , by=list(measure, period)
  )
)
```
The first argument of `part_whole_relation()` is the name of the variable
containing the values. Here, the column `value` from the samplonomy dataset.
The argument `labels` indicates the variable that labels parts and wholes.
Next, we define the label value that indicates a total. Here, a record with
region label `"Samplonia"` indicates a total. Under argument `part` we specify
the labels that have to add up to Samplonia, here the provinces Agria and
Induston. Note that there are more subregions in the dataset, for example the
district of Wheaton (a subregion of Agria). Since we do not specify them, these
are ignored. In the `by` argument we specify that the dataset must be split
into measure and period prior to checking the regional aggregates.


The output is one boolean value per record. For each block, defined by values
of `measure` and `period` either all values are `TRUE`, `FALSE`, or `NA`. The
latter indicates that the aggregate could not be computed because one of the
values is missing, or the computed aggregate could not be compared with the
aggregate in the data because it is missing (either the whole record may be
missing, or the value may be `NA`).
```{r}
out <- confront(samplonomy, rules)
# suppress some columns for brevity
summary(out)[1:7]
```
We can extract the truth values and then inspect the blocks with erroneous values
using standard R functionality.
```{r}
violating(samplonomy, out)
```
Recall that the rule was executed per block defined by `measure` and `period`.
Thus, the result  indicates three errors: one in the block of records defined
by `measure=="gdp"` and `period=="2015"`, also in the blocks defined by
`measure=="export"` and `period==2018Q2` or `period=="2018Q3"`. 

First, it seems that the 2015 GDP of Agria 
is missing from the data set. This turns out indeed to be the case.
```{r}
subset(samplonomy, region=="Agria" & period == "2015" & measure == "gdp")
```
Second, it can be seen that for Induston, there are two export values for
`"2018Q2"` while the export value for `"2018Q3"` is missing.


#### Notes {-}

Specifying (group-wise) aggregates is fairly detailed job in the case of long
data (compare this with 
[balance checks](balance-equalities-and-inequalities)). There are a few things to keep in
mind when using this function.

- The argument `part` is optional. If not specified, every record not matching
  with `whole` will be considered a detail that is to be used to compute the total.
  In the current example this was not possible because besides Agria and Induston,
  we have other subregions.
- In the example we used literal values to specify the keys that define parts
  and wholes. It is possible to, recognize patterns, for example
  any years (4 digits) as a whole and a quarter as a part. See also the next example.
  Supported patterns include regular expressions (shown here) and globbing (see help file).
- It is important that the variables listed in `by` (if any) uniquely specify a
  single aggregate. So here, for each measure and period, the label `"Samplonia"`
  should occur at most once (if it does not occur the result will be `NA`).
- The default way to aggregate is to take the sum. You can specify other ways
  to aggregate by passing an `aggregator` argument. For example `aggregator=mean`.
- By default, the aggregate in the data is compared with the computed aggregate
  up to a tolerance of $10^{-8}$. This tolerance can be set using the `tol`
  argument.  E.g. for integer data you may want to set `tol=0`.

## Aggregates of time series in long format

We are going to check whether quarterly time series add up to the annual time
series.  This is more complicated because of two subtleties.

First there is not one fixed aggregate key, like `"Samplonia"`.  Rather, we
have a key _pattern_. Each total is defined by a period label that consists of
precisely four digits. So rather than recognizing a specific year we want to
recognize that a key represents any year. This can be done using a regular
expression of the form `"^\\d{4}$"`, where the `^` indicates 'start of string',
the `\\d{4}` indicates 'four times a digit' and `$` indicates 'end of string'.

Second, we wish to check annual totals against the sum over quarters for each region and each
measure. However, a value-combination of measure and region does not single out
a single value for `year`. For example, for the Induston export we have the following
annual data.

```{r}
subset(samplonomy, region=="Induston" & freq == "A" & measure=="export")
```
So in fact, we need to do the check _by year_ as well as by measure and region.
Fortunately, in this case it is easy to derive a variable that indicates the year
by selecting the first four characters from `period`.

```{r}
rules <- validator(part_whole_relation(value
  , labels = period
  , whole  = rx("^\\d{4}$")
  , by = list(region, substr(period,1,4), measure) 
  ))
out <- confront(samplonomy, rules)
```
We use `rx("^\\d{4}")` to tell `part_whole_relation` that this string must be
interpreted as a regular expression.  Here, we do not indicate `part` labels
explicitly: by default any record not matching `whole` will be treated as a
detail that must be used to compute the total.

```{r}
errors(out)
# suppress some columns for brevity
summary(out)[1:7]
```

We now get 9 fails and 10 missing values. We can filter out records that
have `NA` (lacking) results.
```{r}
lacking(samplonomy, out)
```
There are two blocks where the annual total could not be compared with 
the sum over quarterly series. The balance value of Crowdon is missing
for `"2014Q1"` as well as the import value of Wheaton for `"2019Q2"`.


Similarly, we can inspect the failing blocks
```{r}
violating(samplonomy, out)
```








