---
title: "Data quality indicators with the validate package"
author: "Mark van der Loo and Edwin de Jonge"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02_Indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE}
library(validate)
```


We assume that the reader went through the first couple of sections
of the [introductory vignette](introduction.html).

### Workflow

In the validate package, an 'indicator' is a rule or function that takes as
input a data set and outputs a number. Indicators are usually
designed to be easily interpretable by domain experts and therefore
depend strongly on the application. In `validate` users are completely free
to define their own indicators. Here is an example of the workflow.

```{r}
i <- indicator(
    mh  = mean(height)
  , mw  = mean(weight)
  , BMI = (weight/2.2046)/(height*0.0254)^2 )
ind <- confront(women, i)
```
In the first statement we define an `indicator` object storing indicator expressions.
Next, we confront a dataset with these indicators. The result is an object of
class `indication`. It prints as follows.
```{r}
ind
```
To study the results, the object can be summarized.
```{r}
summary(ind)
```
Observe that the first two indicators result in a single value (`mh`, `mw`) and 
the third one results in 15 values (`BMI`). The columns `error` and `warning`
indicate wether calculation of the indicators was problematic. 

A specific problem that may occur is when the result of an indicator is non-numeric.
```{r}
i <- indicator(mh = mean(height), a = {"A"})
```
here, the second 'indicator' is an expression that always yields a constant (the character string `"A"`).
```{r}
cf <- confront(women, i)
cf
warnings(cf)
```


### Workflow

