---
title: "Data quality indicators with the validate package"
author: "Mark van der Loo and Edwin de Jonge"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{02_Indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE}
library(validate)
```


We assume that the reader went through the first couple of sections
of the [introductory vignette](introduction.html).

In the validate package, an 'indicator' is a rule or function that takes as
input a data set and outputs a number. Indicators are usually
designed to be easily interpretable by domain experts and therefore
depend strongly on the application. In 'validate' users are free
to specify indicator. By specifing them separate from the programming
workflow, they can be treated as first-class objects: indicator specs can
be maintained, version-controlled, and documented in separate files (just like
validation rules.)

### Workflow

Here is a simple example of the workflow.

```{r}
i <- indicator(
    mh  = mean(height)
  , mw  = mean(weight)
  , BMI = (weight/2.2046)/(height*0.0254)^2 )
ind <- confront(women, i)
```
In the first statement we define an `indicator` object storing indicator expressions.
Next, we confront a dataset with these indicators. The result is an object of
class `indication`. It prints as follows.
```{r}
ind
```
To study the results, the object can be summarized.
```{r}
summary(ind)
```
Observe that the first two indicators result in a single value (`mh`, `mw`) and 
the third one results in 15 values (`BMI`). The columns `error` and `warning`
indicate wether calculation of the indicators was problematic. 

A specific problem that may occur is when the result of an indicator is non-numeric.
```{r}
jj <- indicator(mh = mean(height), a = {"A"})
```
here, the second 'indicator' is an expression that always yields a constant (the character string `"A"`).
```{r}
cf <- confront(women, jj)
cf
warnings(cf)
```


### Getting the values

Values can be obtained with the `values` function, or by converting to a `data.frame`.

We add a unique identifier (this is optional) to make it easier to connect results
with the data.
```{r}
women$id <- letters[1:15]
```
Compute indicators and convert to `data.frame`.
```{r}
ind <- confront(women, i,key="id")
(out <- as.data.frame(ind))
```
Observe that there is no key for indicators `mh` and `mw` since these are
constructed from multiple records.


### Indicators and data.frames

Indicators can be constructed from and coerced to data.frames. To define an indicator you
need to create a data.frame that _at least_ has a character column called `rule`. All
other columns are optional.

```{r}
idf <- data.frame(
  rule = c("mean(height)","sd(height)")
  , label = c("average height", "std.dev height")
  , description = c("basic statistic","fancy statistic")
)
i <- indicator(.data=idf)
i
```

Now, confront with data and merge the results back with rule metadata.
```{r}
quality <- as.data.frame(confront(women, i))
measures <- as.data.frame(i)
merge(quality, measures)
```









