---
title: "Introduction to Validate"
author: "Mark van der Loo Edwin de Jonge"
date: "June 24, 2014"
output: html_document
---
<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Introduction to validate}
-->


## Introduction
The validate package is intended to make checking your data easy, maintainable and reproducible. The validate package
allows you to
- Test your dataset(s) against predefined rules, either in- or cross-dataset
- Store and re-use user-defined indicators, such as summary statistics
- Compare multiple versions of your dataset as it gets processed step by step during for example data cleaning.

We've added plenty of summarizing and visualisations so you can get a quick and attractive look at the results.

### A quick example
Here's an example demonstrating the typical workflow. We'll use the built-in ```women``` dataset.
```{r }
data(women)
summary(women)
```
Validating data is all about checking whether a dataset meets presumtions or expectiations you have about it, and the validate package makes is super easy for you to define those expectations. Let's define a ```validator``` object
that checks whether height and weight are positive.
```{r}
library(validate)
v <- validator(height > 0, weight > 0)
v
```
The validator object has stored the rule and assigned names to them (V0001 and V0002) for future reference. 
To check this, we confront the dataset with the validation rules we've just defined:
```{r}
cf <- confront(v, women)
cf
```
The object ```cf``` contains the result of checking the data in ```women``` against the expectations in ```v```. The 
fact that there are no warnings or errors means that indeed each rule could be evaluated sucessfully (an error would occur for example, if we'd missspell height). Now let's take a look at the acual results.
```{r}
summary(cf)
```
Here, we see that of the 15 records in the ```women``` dataset, 15 pass the test. Consequently, there are no records
that fail the test or yield ```NA```. Now, suppose that we expect that the BMI (weight divided by height squared) of each
person to be below 23. We need to express the weight in kg and the height in meters, so the equation for BMI becomes
$$
BMI = \frac{weight\times0.45359}{(height\times0.0254)^2}
$$
Moreover, we demand that the average BMI is between 22 and 22.5.
Let's create another validator object that first computes the BMI and next test whether the BMI values meet our expectations.
```{r}
v <- validator(
  BMI := (weight*0.45359)/(height*0.0254)^2
  , height > 0
  , weight > 0
  , BMI < 23
  , mean(BMI) > 22 & mean(BMI) < 22.5
)
v
```
Checking is easy as before:
```{r}
cf <- confront(v,women)
summary(cf)
```
We see that of 15 records, 10 meet our demands and that the mean BMI is not within the range we expected.

## Validation rules

### What are validation rules?

Informally, a validation rule states an expectation or presumption that one wishes a data set to satisfy before reliable statistical analyses can take place. It is common to distinguish between the following types of validation rules.

- **In-variable validation rules.** These concern basic range checks, for example: is the weight variable nonnegative?
- **Cross-variable, in-record validation.** These rules check inter-variable relations. The test on BMI-values in the quick
example above is an example of cross-variable validation. 
- **Cross-record validation.** These rules are characterised by the fact that theire evaluation depends on more than a single record. For example, one might expect that no _salary_ in a column of salaries exceeds ten times
the median salary of the column. The test on the mean BMI-value in the quick example above is another demonstration of cross-record validation. 
- **Cross-dataset validation.** Compare (statistics of) multiple datasets. For example, one may expect that this year's revenue is within 10% of last year's.

The validate package offers facilities for all these types of validation rules.

Technically, a validation rule is an expression (or function, the difference between the two is not very big) that takes some data and returns either ```TRUE```, ```FALSE``` or ```NA```. The latter occurs only when the dataset contains too many missing values to compute the result of the expression. In the validate package, a validation rule is the subset of R-expressions whose final result is determined by any of the following operators or functions.
```{r,echo=FALSE}
s <- validate:::SYNTAX$validationsymbols
s[!s%in% c(':=',':')]
```
Here, functions ```V``` and ```L``` will be discussed below in the section on [impact and severity](#impactseverity).

### Storing and retrieving validation rules

#### From the commandline


#### From file
```{r,eval=FALSE}
validator(files= "myrules.txt")
```

#### Multiple variables, the same rules

#### Impact and Severity
<a name='impactseverity'></a>


### Properties of the validator object

### Analysing confrontations

## Indicators

### What are indicators?

### Counting missing values

### Properties of the indicator object

### Analysing confrontations


## Comparing datasets

