---
title: "Introduction to Validate"
author:
- Mark van der Loo
- Edwin de Jonge
date: "June 24, 2014"
output: html_document
---
<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Introduction to validate}
-->


## Introduction
The validate package is intended to make checking your data easy, maintainable and reproducible. Here's an example demonstrating the
typical workflow.

We'll use the built-in ```women``` dataset
```{r }
data(women)
```

1. Define the rules you want your data set or sets to obey by creating
a ```validator``` object.
```{r}
library(validate)
v <-validator(height > 0, weight > 0,mean(heigt)>66)
```

2. Confront your data with the rules
```{r}
cf <- confront(v, women)
```

3. Analyse the results
```{r}
summary(cf)
values(cf)
```
## A simple example






- define validation rules: `validator` and `indicator`
- confront with data `confront`, `validatorValue`, `indicatorValue`
- use result of confrontation `compare`

## Defining validation rules

## Definining indicators

## Confronting with data

## Comparing