---
title: "Introduction to Validate"
author: "Mark van der Loo and Edwin de Jonge"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{01_Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE}
library(validate)
```

### Introduction

The validate package is intended to make checking your data easy, maintainable
and reproducible. The package allows you to

- Test your data set(s) against predefined rules, either in- or cross-data set
- Import and export rule sets from structured or free-format files
- Investigate and visualize the results of a data validation step
- Perform basic rule maintenance tasks
- Define and maintain data quality indicators separately from the data


There are a few nouns related to the infrastructure offered by validate,

- data set: your data
- `validator` an object representing a set of rules your data must satisfy
- `indicator` an object representing a set of numerical quality indicators
- `confrontation` an object representing the results of confronting data with rules or
  quality indicators.

There is also a single verb, namely

- `confront`: evaluate the validation rules or quality indicator in the context of one or more data sets.

This vignette demonstrates how to apply validation rules. A [second
vignette](indicators.html) introduces quality indicator rules.


### A quick example
Here's an example demonstrating the typical workflow. We'll use the built-in
```women``` data set (average heights and weights for American women aged
30-39).
```{r }
data(women)
summary(women)
```
Validating data is all about checking whether a data set meets presumptions or
expectations you have about it, and the validate package makes it easy
for you to define those expectations. Let's do a quick check on variables
in the `women` data set.
```{r}
library(validate)
cf <- check_that(women, height > 0, weight > 0, height/weight > 0.5)
summary(cf)
```
`check_that` returns an object containing all sorts of information on the validation results.
The easiest way to check the results is with `summary`, which returns a `data.frame` with
the following basic information:

- How many data items were checked against each rule
- How many items passed, failed or resulted in `NA`
- Whether the check resulted in an error (could not be performed) or gave an error
- The expression that was actually evaluated to perform the check.

If you're a fan of the pipe-operator provided by the [magrittr](http://www.r-pkg.org/pkg/magrittr),
the above statement can also be performed as follows.
```{r eval=FALSE}
women %>% check_that(height > 0, weight > 0, height/weight > 0.5) %>% summary()
```

The same information can be summarized graphically as well.
```{r}
barplot(cf,main="Checks on the women data set")
```


### Validator objects
Validator objects are used to store, investigate and manipulate rule sets.
```{r}
v <- validator(height > 0, weight > 0, height/weight > 0)
v
```
The validator object has stored the rule and assigned names to them for future
reference.  To check this, we confront the data set with the validation rules
we've just defined:
```{r}
cf <- confront(women,v)
cf
```
The object ```cf``` contains the result of checking the data in ```women```
against the expectations in ```v```. The fact that there are no warnings or
errors means that indeed each rule could be evaluated successfully (an error
would occur for example, if we'd misspell height). Now let's take a look at
the actual results.
```{r}
summary(cf)
```

Now, suppose that we expect that the BMI (weight divided by height squared) of
each item to be below 23. We need to express the weight in kg and the height
in meters, so the equation for BMI becomes
$$
BMI = \frac{weight\times0.45359}{(height\times0.0254)^2}
$$
Moreover, assume that we suspect that the average BMI is between 22 and 22.5. 
Let's create another validator object that first computes the BMI and next tests
whether the BMI values conform to our suspicion.
```{r}
v <- validator(
  BMI := (weight*0.45359)/(height*0.0254)^2
  , height > 0
  , weight > 0
  , BMI < 23
  , mean(BMI) > 22 & mean(BMI) < 22.5
)
v
```
Checking is easy as before:
```{r}
cf <- confront(women,v)
summary(cf)
```
Observe that the expressions for validation are now manipulated: everywhere where `BMI` was used, it was replaced with the computation defined before.

### Conversion from and to `data.frames`

Validator objects can be read from and converted to `data.frames`. To create
a validator object, at least a character column named `rule` is necessary.
```{r}
df <- data.frame(
  rule = c("height>0","weight>0","height/weight>0.5")
  , label = c("height positive","weight positive","ratio limit")
)
v <- validator(.data=df)
v
```
Now confront with the data and merge back with rule metadata. 
```{r}
cf <- confront(women, v)
quality <- as.data.frame(cf)
measure <- as.data.frame(v)
head( merge(quality, measure) )
```

Or, merge with the summarized results. The result of `summary` is just a `data.frame` (tidy, isn't it?).
```{r}
merge(summary(cf),measure)
```



### Confrontation objects
The outcome of confronting a validator object with a data set is an object of class
`confrontation`. There are several ways to extract information from a `confrontation` object.

- `summary`: summarize output; returns a `data.frame`
- `aggregate`: aggregate validation in several ways
- `sort` : aggregate and sort in several ways
- `values`: Get the values in an array, or a list of arrays if rules have different output dimension structure
- `errors`: Retrieve error messages caught during the confrontation
- `warnings`: Retrieve warning messages caught during the confrontation.

By default aggregates are produced by rule.
```{r}
cf <- check_that(women, height>0, weight>0,height/weight < 0.5)
aggregate(cf) 
```
To aggregate by record, use `by='record'`
```{r}
head(aggregate(cf,by='record'))
```
Aggregated results can be automatically sorted, so records with the most violations or
rules that are violated most sort higher.
```{r}
# rules with most violations sorting first:
sort(cf)
```

Confrontation objects can be subsetted with single bracket operators (like vectors), to 
obtain a sub-object pertaining only to the selected rules.
```
summary(cf[c(1,3)])
```


### Confrontation options
By default, all errors and warnings are caught when validation rules are confronted with data. This can be switched off by setting the `raise` option to `"errors"` or `"all"`, e.g. here, the error is caught,
```{r}
v <- validator(hite > 0, weight>0)
summary(confront(women, v))
```
while here it is raised immediately.

```{r eval=FALSE}
# this gives an error
confront(women, v, raise='all')
```


Linear equalities form an important class of validation rules. To prevent equalities to be strictly tested, there is an option called `len.eq.eps` (with default value $10^{-8}$) that allows one to add some slack to these tests. The amount of slack is intended to prevent false negatives (unneccesary failures) caused by machine rounding. If you want to check whether a sum-rule is satisfied to within one or two units of measurement, it is cleaner to define two inequalities for that.





### Metadata and investigating validator objects
Validator objects store a set of rules, optionally with some metadata per rule.
Currently, the following functions can be used to get or set metadata:

- `origin` : Where was a rule defined?
- `names` : The name per rule
- `created` : when were the rules created?
- `label` : Short description of the rule
- `description`: Long description of the rule
- `meta`: Set or get generic metadata

For example, names can be set from the command line when defining a validator object.
```{r}
v <- validator(rat = height/weight > 0.5, htest=height>0, wtest=weight > 0)
names(v)
```
Also try
```{r}
names(v)[1] <- "ratio"
v
```

It is also possible to add generic key-value pairs as metadata. Getting and setting follows the usual
recycling rules of R.
```{r}
# add 'foo' to the first rule:
meta(v[1],"foo") <- 1
# Add 'bar' to all rules
meta(v,"bar") <- "baz"
```
Metadata can be made visible by selecting a single rule:
```{r}
v[[1]]
```
Or by extracting it to a `data.frame`
```{r}
meta(v)
```


Some general information is obtained with `summary`,
```{r}
summary(v)
```
Here, some properties per _block_ of rules is given. Two rules occur in the same
block if when they share a variable. In this case, all rules occur in the same
block.

and the number of rules can be requested with `length`
```
length(v)
```
With `variables`, the variables occurring per rule, or over all the rules can be requested.
```{r}
variables(v)
variables(v,as="matrix")
```
Validator objects can be subsetted as if they were lists using the single and double bracket operators.
```{r}
v[c(1,3)]
v['ratio','wtest']
```

The double bracket can be used to inspect a single rule
```{r}
v[[1]]
```


### Validator objects and confrontation objects are reference objects
As simple as that. If you do
```
w <- v
```
for a validator object `v`, then `w` just points to the same physical object as `v`.
To make an actual copy, you can select everything.
```
w <- v[]
```





