---
title: "Rules in text files"
author: "Mark van der Loo and Edwin de Jonge"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: united
vignette: >
  %\VignetteIndexEntry{Rules in text files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### Introduction

For editing, reuse, maintaining and sharing of validation rules it is convenient
to define them in text files. In this vignette we demonstrate how rules can be imported
from or exported to text files.

### Importing from free form text files
The easiest way to define rules is by storing them in a free-form text file. For example,
create a file called `ex-1.txt`
```
# content of ex-1.txt

# Some checks on the 'women' dataset. 
height > 0

weight > 0

mean(height/weight) < 0.5
```
We can read the rules using
```{r,eval=FALSE}
v <- validator(.file="ex-1.txt")
```
Note that the `#` is used to start comment-lines, just like in regular R syntax. Validation rules may span several lines. The only restriction is that rules are stated in valid syntax and can be recognized by validate as validating (basically, it should result in a `logical`).


### Importing from yaml files
To set options and metadata for the rules, the well-known [yaml](www.yaml.org) (yaml ain't markup language) format is used. Yaml is a human-readable way to define (nested) structures. Here is an example
of a yaml-based rule definition file.
```
# content of ex-2.yaml
rules:
- 
  expr: height > 0
  name: height
  label: height positivity
  description: |
   According to the latest research, the average height of American women
   must be positive. 
-
  expr: weight > 0
  name: weight
  label: weight positivity
  description: |
    By definition, weight must be positive.
```
This file can be read in the same way.
```
v <- validator(.file='rules.yaml')
```
There are a few things to note here:

- Defining a set of rules with metadata starts with the `rules:` directive.
- Every dash starts a new rule.
- Indentation matters. In `yaml` indentation _must be spaces_ and _not tabs_. In the example,
there are two spaces of indentation before a key name (`expr`, `name`, and so on). For long rules
or pieces of text, you can add a `|` or `>` after the colon and add a multiline string. You need
to start on a new line and add an extra indentation.
- Any entry starting with an exclamation mark (```!```) must be enquoted since the exclamation
mark has special meaning in YAML.

The latter remark means that this is wrong:
```
# READING THIS FILE FAILS
rules:
- 
  expr: !is.na(height) 
```

and this is ok:
```
# READING THIS FILE SUCCEEDS
rules:
- 
  expr: '!is.na(height)' 
```

Rules exported with ```export_yaml``` are enquoted by default.






### Mixing free form and yaml
This is possible. Just separate free-form sections from structured sections with
three dashes on a single line.
```
# content of ex-3.yaml
rules:
- 
  expr: height > 0
  name: height
  label: height positivity
  description: |
   According to the latest research, the average height of American women
   must be positive. 
-
  expr: weight > 0
  name: weight
  label: weight positivity
  description: |
    By definition, weight must be positive.
---
# free form starts here

# we expect the following mean ratio
mean(height/weight) < 0.5

# we expect a high correlation
cor(height,weight) > 0.99
```

### Setting options
This can be done at the beginning of your file. Start and end the options section with three dashes
on a line to start the (free form or structured) rule section.
```
---
options:
  raise: errors
---
height > 0
```
The options you set here will be part of the `validator` object, that is created once you read in the file. The options are valid for every confrontation you use this validator for, unless they are overwritten during the call to `confront`.

### Including other rule files
This is useful, for example when you have a general rule set, that applies to all your data
files, and some rules that apply only in specific cases. The file with specific cases can include
one or more rule files in that case. Files should be included in the same section as the options.
```
---
include:
  - petes_rules.yaml
  - nancys_rules.yaml
options:
  raise: errors
---
# start rule definitions here
```

### Exporting validator objects
There are two ways to do that. You can either write to a `yaml` file immediately as follows
```{r, eval=FALSE}
v <- validator(height>0, weight> 0)
export_yaml(v,file="my_rules.yaml")
```
or you can get the `yaml` text string using `as_yaml`
```
cat(as_yaml(v))
```
















