# Introduction to validate {#intro}

> Data validation is an activity where one checks whether a set of values
> or value combinations satisfies a predefined expectation.


The validate package is intended to make checking your data easy, maintainable, and reproducible.
It does this by allowing you to

- test data against a reusable set of rules:
- investigate, summarise, and visualise data validation results;
- import and export rule sets from and to various formats;
- filter, select and otherwise manipulate data validation rules';
- investigate, summarise, and visualise rule sets.

For advanced rule manipulation there is the
[validatetools](https://cran.r-project.org/validatetools) package.



## A quick example

Here's an example demonstrating the typical workflow. We'll use the built-in
`cars` data set, which contains 50 cases of speed and stopping distances of
cars. 
```{r }
data(cars)
head(cars, 3)
```
Validating data is all about checking whether a data set meets presumptions or
expectations you have about it, and the validate package makes it easy for you
to define those expectations. Let's do a quick check on variables in the
`cars` data set. We first load the package, and create a list
of data quality demands with the `validator()` function.
```{r}
library(validate)
rules <- validator(speed >= 0, dist >= 0, speed/dist <= 1.5, cor(speed, dist)>=0.2)
```
Here, the first three rules are record-wise checks: each record will yield one
answer. In the last rule we check whether speed and distance are positively
correlated this will yield a single `TRUE` or `FALSE` for the whole data set.
We now confront the data with those rules and save the output into a variable
called `out`.
```{r}
out   <- confront(cars, rules)
```
The easiest way to check the results is with `summary()`. 
```{r}
summary(out)
```
This returns a data frame with one line of information for each rule `V1`,
`V2`, `V3` and `V4`. To be precise:

- How many data items were checked against each rule.
- How many items passed, failed or resulted in `NA`.
- Whether the check resulted in an error (could not be performed) or gave an warning.
- The expression that was actually evaluated to perform the check.

The same information can be summarized graphically.
```{r,fig.height=7, fig.width=7}
plot(out)
```

In this plot each horizontal bar indicates the percentage of Failing, Passing,
and Missing cases. The table in the legend lists the total number of Fails,
Passes and Missings, summed over all checks. Here, we have 4 rules. The first
three rules yield 50 results each, while the last rule yields a single result.
Hence there are 151 validation results in total.



Using the function `violating` we can select the records that violate one or
more rules. We select only the first three results because the last rule can
not be interpreted record by record.
```{r}
violating(cars, out[1:3])
```

We can extract all individual resuls results using for example `as.data.frame`.
```{r}
df_out <- as.data.frame(out)
head(df_out, 3)
```
We see that in record `1`, rule `V1`, was satisfied (the result is `TRUE`), and
that validate left a bit of slack when executing the rule, to avoid false
negatives caused machine rounding issues.


Summarizing, the basic workflow in `validate` is to create a rule set, confront
a data set with the rules in the rule set, and than analyze or use the results
further. To understand which checks you can perform with `validate` you only
need to remember the following.

> Any R expression that results in a `logical` is accepted by `validate` as a
> validation rule.


You are now ready to start validating your data, and navigate Chapters
\@ref(sect:availableunique)-\@ref(sect:statisticalchecks) to learn how to
define specific types of checks. Chapter~\@ref(sect:work), discusses more
details about working with `validate`.



