# Variable checks {#sect-varlevelchecks}

```{r, include=FALSE}
source("chunk_opts.R")
```


Variable checks are checks that can be performed on a field-by-field basis. An
example is checking that a variable called `Age` is nonnegative, or of integer
type. Variable checks are among the simplest checks.


**Data**

In this section we will use the `SBS2000` dataset, that is included with `validate`.

```{r}
library(validate)
data(SBS2000)
head(SBS2000, 3)
```

See `?SBS2000` for a description.


## Variable type

In `R`, one can test the type of a variable using built-in functions such as
`is.numeric` or `is.character`. 
```{r}
is.character("hihi")
is.character(3)
```

In `validate`, any function starting with `is.` ('is' followed by a dot) is
considered a validation function. 

```{r}
rules <- validator(
    is.character(size)
  , is.numeric(turnover)
)
out <- confront(SBS2000, rules)
summary(out)
```
We see that each rule checks a single item, namely one column of data. The
first rule is violated (it is in fact a `factor` variable). The second rule
is satisfied.

## Missingness {#sect-missingness}

Use R's standard `is.na()` to check missing items in individual variables. Negate
it to check that values are available.

```{r}
rule <- validator(
   !is.na(turnover)
 , !is.na(other.rev)
 , !is.na(profit)
)
out <- confront(SBS2000, rule)
summary(out)
```

We see that in `r summary(out)$fails[1]` cases the variable `turnover` is missing,
while `other.rev` and `profit` are missing respectively in `r summary(out)$fails[2]`
and `r summary(out)$fails[3]` occasions.

To demand that all items must be present or absent for a certain variable,
use R's  quantifiers: `any()` or `all()`, possibly negated.
```{r}
rules <- validator( 
    !any(is.na(incl.prob))
    , all(is.na(vat)) )
out <- confront(SBS2000, rules)
summary(out) 
```


- To check whether records or parts thereof are completed, see \@ref(sect-iscomplete).
- To check whether records are available at all, see \@ref(sect-completeness).


## Field length

The number of characters in text fields can be tested using either R's standard
`nchar()` function, or with the convenience function `field_length`.

```{r}
rules <- validator(
   nchar(as.character(size)) >= 2
 , field_length(id, n=5)
 , field_length(size, min=2, max=3)
)
out <- confront(SBS2000, rules)
summary(out)
```
One advantage of `check_field_length` is that its argument is converted to
character (recall that `size` is a `factor` variable). The function
`field_length` can be used to either test for exact field lengths or to
check whether the number of characters is within a certain range.

The field length is measured as the number of [code
points](https://en.wikipedia.org/wiki/Code_point). Use `type="width"` to
measure the printed width (nr of columns) or `type="bytes"` to count the number
of bytes. 




## Format of numeric fields

For numbers that are stored in `character` type, there is a convenience
function called `number_format()` that accepts a variable name
and a format specification.

```{r}
dat <- data.frame(x = c("2.54","2.66","8.142","23.53"))
```
To check that the numbers are formatted with one figure before, and two figures
after the decimal point, we perform the following check.
```{r}
rule <- validator( number_format(x, format="d.dd"))
values(confront(dat, rule))
```
Here, the specification `format="d.dd"` describes the allowed numeric formats.
In this specification the `"d"` stands for a digit, any other character except
the asterisk (`*`) stands for itself. The asterisk is interpreted as 'zero or
more digits'. Here are some examples of how to define number formats.

|format       | match                         | non-match                           |
|-------------|-------------------------------|-------------------------------------|
|`0.dddd`     | `"0.4321"`                    | `"0.123"`,`"1.4563"`                |
|`d.ddEdd`    | `"3.14E00"`                   | `"31.14E00"`                        |
|`d.*Edd`     | `"0.314E01"`,`"3.1415297E00"` | `"3.1415230"`                       |
|`d.dd*`      | `"1.23"`, `"1.234"`,$\ldots$  | `"1.2"`                             |

The last example shows how to check for a minimal number of digits behind the decimal
point.


There are special arguments to check the number of decimal figures 
after the decimal separator. 
```{r}
x <- c("12.123","123.12345")
number_format(x, min_dig=4)
number_format(x, max_dig=3)
number_format(x, min_dig=2, max_dig=4)
number_format(x, min_dig=2, max_dig=10)
# specify the decimal separator.
number_format("12,123", min_dig=2, dec=",")
```
The arguments `min_dig`, `max_dig` and `dec` are ignored when `format` is
specified.

This function is convenient only for fairly simple number formats. Generic
pattern matching in strings is discussed in the next section.



## General field format

A simple way to check for more general format is to use [globbing
patterns](https://en.wikipedia.org/wiki/Glob_(programming)). In such patterns,
the asterisk wildcard character (`*`) is interpreted as 'zero or more
characters' and the question mark (`?`) is interpreted as 'any character'.

For example, to check that the `id` variable in `SBS2000` starts with `"RET"`,
and that the `size` variable has consists of `"sc"` followed by precisely one
character, we can do the following.
```{r}
rule <- validator(field_format(id, "RET*")
                , field_format(size, "sc?" ))
out  <- confront(SBS2000, rule)
summary(out)
```
Here, the globbing pattern `"RET*"` is understood as 'a string starting with
`"RET"`, followed by zero or more characters. The pattern `"sc?"` means 'a
string starting with `"sc"`, followed by a single character.


The most general way to check whether a field conforms to a pattern is to use a
[regular expression](https://en.wikipedia.org/wiki/Regular_expression). The
treatment of regular expressions is out of scope for this book, but we will
give a few examples. A good introduction to regular expressions is given by

> J. Friedl (2006) _Mastering regular expressions._ O'Reilley Media.


In `validate` one can use `grepl` or `field_format`, with the argument `type="regex"`

```{r}
rule <- validator(
          grepl("^sc[0-9]$", size)
        , field_format(id, "^RET\\d{2}$" , type="regex") )
summary(confront(SBS2000, rule))
```

Here, the expression `"^sc[0-9]$"` is a regular expression that should be read
as: the string starts (`"^"`) with `"sc"`, is followed by a number between 0
and 9 (`"[0-9]"`) and then ends (`"$"`). The regular expression `"^RET\\{d}2"`
indicates that a string must start (`"^"`) with `"RET"`, followed by two
digits (`"\\d{2}"`), after which the string must end (`"$"`).


Globbing patterns are easier to develop and easier to understand  than regular
expressions, while regular expressions offer far more flexibility but are
harder to read. Complex and long regular expressions may have subtle matching
behaviour that is not immediately obvious to inexperienced users. It is
therefore advisable to test regular expressions with a a small dataset
representing realistic cases that contains both matches and non-matches. As a
rule of thumb we would advise to use globbing patterns unless those offer
insufficient flexibility.



## Numeric ranges

Numerical variables may have natural limits from below and/or
above. For one-sided ranges, you can use the standard comparison operators.

```{r}
rules <- validator(TO = turnover >= 0
                 , TC = total.costs >= 0)
```

If a variable is bounded both from above and below one can use two rules,
or use the convenience function `in_range`.

```{r}
rules <- rules + 
  validator(PR = in_range(incl.prob, min=0, max=1))
```
By default, `in_range` includes the boundaries of the range, so the above rule
is equivalent to `incl.prob >= 0` and `incl.prob <= 1`. 

```{r}
out <- confront(SBS2000, rules, lin.ineq.eps=0)
```
Here we set `lin.ineq.eps=0` to keep `validate` from building in a 
margin for machine rounding errors.

```{r}
summary(out)
```

For numeric ranges it is often a better idea to work with inclusive
inequalities ($\leq$, $\geq$) than with strict inequalities ($<$, $>$).  Take
as an example the strict inequality demand `income > 0`. This means that _any_
income larger than zero is acceptable, including numbers such as $0.01$,
$0.000001$ and $10^{-\textrm{Googol}}$. In practice there is almost always a
natural minimal acceptable value that is usually dictated by the unit of
measurement. For example, if we measure income in whole Euros, a better demand
would be `income >= 1`.


## Ranges for times and periods

For objects of class `Date` and objects of class `POSIXct` one can use comparison
operators and `in_range` in the same way as for numerical data. The `in_range` function
has a convenience feature for period data that is coded in character data, as in 
`"2018Q1"` for quarterly data. 

We first generate some example data.
```{r}
period = sprintf("2018Q%d", 1:4)
period
```

The `in_range` function is capable of recognizing certain date or period
formats.
```{r}
in_range(period, min="2017Q2", max = "2018Q2")
```
It is possible to specify your own date-time format using `strftime` notation.
See `?in_range` and `?strptime` for specifications.



## Code lists

A code list is a set of values that a variable is allowed to assume. For small
code lists, one can use the `%in%` operator.

```{r}
rule <- validator(size %in% c("sc0","sc1","sc2","sc3"))
out  <- confront(SBS2000, rule)
summary(out)
```
Notice that `validate` replaces `%in%` with `%vin%`. The reason is that `%vin%` has more consistent
behavior in the case of missing data. In particular,
```{r}
c(1, 3, NA) %in% c(1,2)
c(1, 3, NA) %vin% c(1,2)
```

For longer code lists it is convenient to refer to an externally provided list.
There are two ways of doing this: reading the list in the right-hand-size of `%in%`,
or passing a code list to `confront` as reference data.

Suppose we have a file called `codelist.csv` with a column `code`. We can define
a rule as follows.

```{r}
rule <- validator(
  x %in% read.csv("codelist.csv")$code
)
## Or, equivalently
rule <- validator(
  valid_codes := read.csv("codelist.csv")$code
  , x %in% valid_codes
)

```
The disadvantage is that the rule now depends on a path that may or may not be
available at runtime.

The second option is to assume that a variable, say `valid_codes` exists at
runtime, and pass this with `confront`.
```{r}
codelist <- c("sc0","sc1","sc2","sc3")
rule <- validator(size %in% valid_codes)
# pass the codelist
out <- confront(SBS2000, rule
              , ref=list(valid_codes=codelist))
summary(out)
```
This way, (very) large code lists can be used, but note that it does
require a 'contract' between variable names used in the rule set
and variables passed as reference data.








