# Multivariate checks


```{r, include=FALSE}
source("chunk_opts.R")
```

In this Chapter we treat tests that involve relationships between variables.


**Data**

In this Chapter we will use the `SBS2000` dataset that comes with `validate`.

```{r}
library(validate)
data(SBS2000)
head(SBS2000, 3)

```

## Completeness of records {#sect:iscomplete}

The functions `is_complete()` and `all_complete()` are convenience functions
that test for missing values or combinations thereof in records.

```{r}
rules <- validator(
          is_complete(id)
        , is_complete(id, turnover)
        , is_complete(id, turnover, profit )
        , all_complete(id)
)
out <- confront(SBS2000, rules)
# suppress last column for brevity
summary(out)[1:7]
```
Here, the first rule checks for missing data in the `id` variable, the second
rule checks whether subrecords with `id` and `turnover` are complete, and the
third rule checks whether subrecords with `id`, `turnover` and `profit` are
complete.  The output is one logical value (`TRUE` or `FALSE`) for each record.

The fourth rule tests whether _all_ values are present in the `id` column, and
it results in a single `TRUE` or `FALSE`. 

- To test for missing values in individual variables, see also \@ref(sect:missingness).
- To check whether records are available at all, see \@ref(sect:completeness).


## Balance equalities and inequalities 

Balance restrictions occur for example in economic microdata, where financial
balances must be met.
```{r}
rules <- validator(
    total.rev - profit == total.costs
  , turnover + other.rev == total.rev
  , profit <= 0.6*total.rev
)

out <- confront(SBS2000, rules)
summary(out)
```
Here, the first rule checks a balance between income, costs, and profit; the
second rule checks a sub-balance, and the third rule is a plausibility check
where we do not expect profit to exceed 60 per cent of the total revenue.

Observe that the expressions have been altered by `validate` to account for
possible machine rounding differences. Rather than testing whether variable $x$
equals variable $y$, `validate` will check $|x-y|\leq \epsilon$, where the
default value of $\epsilon$ is $10^{-8}$. The value of this tolerance can be
controlled for linear equalities and inequalities using respectively
`lin.eq.eps` and `lin.ineq.eps`.

```{r}
out <- confront(SBS2000, rules, lin.ineq.eps=0, lin.eq.eps=0.01)
summary(out)
```

See \@ref(sect:options) for more information on setting and resetting options.


## Conditional restrictions

Conditional restrictions are all about demanding certain value combinations.
In the following example we check that a business with staff also has staff
costs.
```{r}
rule <- validator(if (staff >= 1) staff.costs >= 1)
out  <- confront(SBS2000, rule)
summary(out)
```
Here, combinations where there is a positive number of staff must be
accompanied with a positive staff cost. 

Validate translates the rule `if ( P ) Q` to an expression of the form `!P |
Q`. The reason for this is that the latter can be evaluated faster
(vectorised).

The results are to be interpreted as follows. For each record, `validate`  will
check that cases where `staff>=1` are accompanied by `staff.costs >= 1`. In
cases where this test results in `FALSE` this means that either the staff
number is too high, or the staff costs are too low. To be precise, the results
of a conditional restriction match those of an implication in first-order
logic as shown in the truth table below.

$$
\begin{array}{ll|c}
P & Q & P\Rightarrow Q\\
\hline
T & T & T\\
T & F & F\\
F & T & T\\
F & F & F\\
\end{array}
$$



## Forbidden value combinations

In some cases it is more convenient to have a list of forbidden (key) value
combinations than specifying such combinations individually.  The function
`does_not_contain()` supports such situations.


As an example, let's first create some transaction data.
```{r}
transactions <- data.frame(
   sender   = c("S21", "X34", "S45","Z22")
 , receiver = c("FG0", "FG2", "DF1","KK2")
 , value    = sample(70:100,4)
)
```
We assume that it is not possible for senders with codes starting with an `"S"`
to send something to receivers starting with `FG`. A convenient way to encode
such demands is to use 
[globbing patterns](https://en.wikipedia.org/wiki/Glob_(programming)).
We create a data frame that lists forbidden combinations (here: one combination
of two key patterns).

```{r}
forbidden <- data.frame(sender="S*",receiver = "FG*")
```
Note that the column names of this data frame correspond to the columns in the transactions
data frame. We are now ready to check our transactions data frame.
```{r}
rule <- validator(does_not_contain(glob(forbidden_keys)))
out <- confront(transactions, rule, ref=list(forbidden_keys=forbidden))
## Suppress columns for brevity
summary(out)[1:7]
```
Observe that we use `glob(forbidden_keys)` to tell `does_not_contain` that the
key combinations in the `forbidden_keys` must be interpreted as globbing
patterns.

The records containing forbidden keys can be selected as follows.
```{r}
violating(transactions, out)
```

It is also possible to use regular expression patterns, by labeling the
forbidden key set with `rx()`. If no labeling is used, the key sets are
interpreted as string literals.



