# Statistical checks {#sect:statisticalchecks}

Statistical checks involve group properties such as the means of columns. These
characteristics can be checked for whole columns or grouped by one or more
categorical variables. It is also possible to use group-wise computed
statistics in validation rules.  For example if you want to compare individual
values with a mean within a group.

For long-form data it is possible to compare aggregate values with underlying
details. For example to test whether quarterly time series add up to annual
totals. It is also possible to check properties of groups, for example whether
in every household (a group of persons) there is exactly one head of household.

### Data and packages {-}

In this Chapter we will use the `SBS2000` dataset that comes with `validate`.

```{r}
library(validate)
data(SBS2000)
head(SBS2000, 3)
```

We shall also use the `samplonomy` dataset that also comes with `validate`. See also
\@ref(long-data).

```{r}
data(samplonomy)
head(samplonomy, 3)
```


## Statistical and groupwise characteristics {#sect:groupwise}

Any R expression that ultimately is an equality or inequality check is
interpreted as a validation rule by validate. This means that any statistical
calculation can be input to a rule.

Here we check the mean profit and correlation coefficient between profit and
turnover.
```{r}
rule <- validator(
    mean(profit, na.rm=TRUIE) >= 1
  , cor(turnover, staff, use="pairwise.complete.obs") > 0
)
out <- confront(SBS2000, rule)
# suppress some columns for brevity
summary(out)[1:7]
```

There are a few helper functions to compute group-wise statistics, and to make
comparing values with group aggregates possible.

For example, here we check whether each turnover is less than ten times
the group-wise median.
```{r}
rule <- validator(
  turnover <= 10*do_by(turnover, by=size, fun=median, na.rm=TRUE)
)
out <- confront(SBS2000, rule)
# suppress some columns for brevity
summary(out)[1:7]
```
Here, in the right-hand side of the rule the group-wise median of turnover is
computed. The function `do_by` is very similar to functions such as `tapply`
in base R. The difference is that `do_by` works on vectors only (not on data
frames) and always repeats the values of `fun` so that the length of the output is
equal to the length of the input.
```{r}
medians <- with(SBS2000, do_by(turnover, by=size, fun=median, na.rm=TRUE))
head(data.frame(size = SBS2000$size, median=medians))
```
There are also some convenience functions, including `sum_by`, `mean_by`, `min_by`, and `max_by`.

## Group properties

In this section, we group data by one or more categorical variables and
check for each group whether a rule is satisfied. In particular we are
going to check whether each household in a small dataset has a unique 
'head of household'.

We first create some data with household id (`hhid`) a person id (`person`) and
that person's role in the household (`hhrole`).
```{r}
d <- data.frame(
   hhid   = c(1,  1,  2,  1,  2,  2,  3 )
 , person = c(1,  2,  3,  4,  5,  6,  7 )
 , hhrole = c("h","h","m","m","h","m","m")
)
d
```
With `exists_one()` we can check that there is exactly one person
with the role `"h"` (head) in each household, by grouping on household id.
```{r}
rule <- validator(exists_one(hhrole == "h", by=hhid))
out <- confront(d, rule)
# suppress some columns for brevity
summary(out)
```
We can inspect the results by selecting the violating record groups.
```{r}
violating(d, out)
```
We see that household 1 has two heads of household, while household 3 has no head 
of household. 

To test whether _at least one_ head of household exists, one can use
`exists_any`:

```{r}
violating(d, validator(exists_any(hhrole=="h",by=hhid) ))
```


In the following example we check whether there is exactly one region called Samplonia
for each period and each measure in the `samplonomy` dataset.

```{r}
rule <- validator(exists_one(region=="Samplonia", by=list(period, measure)))
```
The first argument of `exists_one()` is a rule that has to be checked in every group
indicated by the `by` argument. The output is a logical vector with an element for
each record in the dataset under scrutiny. If a group of data fails the test, each record
in that group is indicated as wrong (`FALSE`).

```{r}
out <- confront(samplonomy, rule)
# suppress some columns for brevity
summary(out)[1:7]
```
Here, there are no groups that violate this assumption. 
```{r}
violating(samplonomy, out)
```

## Code hierarchies and aggregation

Classifications and ontologies often have a hierarchical structure.
A well-known example is the 
[NACE](https://en.wikipedia.org/wiki/Statistical_Classification_of_Economic_Activities_in_the_European_Community) classification of economic activities. In the NACE classification, 
the economy is divided into 10 basic types of activities such as 'Agriculture'
or 'Mining and Quarrying', and each activity is again divided into subclasses,
such as 'Growing of rice' and 'Growing of Grapes' under 'Agriculture'. The
subdividing can go on for several levels. For statistics that describe an
economy according to the NACE classification, it is desirable that the
statistics of subclasses add up to their parent classes. This is what the
function 'hierarchy' does in 'validate'.

The `validate` package comes with a version of the NACE classification
(Revision 2, 2008) so we will use that as an example.
```{r}
data(nace_rev2)
head(nace_rev2[1:4])
```
The second and third column contain the necessary information: they list the
parent for each NACE code (where each parent is also a NACE code). To demonstrate
how `hierarchy()` works, we first create some example data.
```{r}
dat <- data.frame(
        nace   = c("01","01.1","01.11","01.12", "01.2")
      , volume = c(100 ,70    , 30    ,40     , 25    )
     )
dat
```
We see that the volumes for subclasses `"01.11"` and `"01.12"` add up to
`"01.1"` ( $30+40=70$ ). However, the volumes for `"01.1"` and `"01.2"` do not
add up to the volume for `"01"` ($70+25\not=100$). The `hierarchy()` function
checks all these relations.


Before using `hierarchy` in the setting of a `validator` object, we can examine
it directly.
```{r}
dat$check <- hierarchy(dat$volume, dat$nace, nace_rev2[3:4])
dat
```
We see that `hierarchy()` returns a `logical` vector with one element for each
record in the data. Each record that is involved in one or more aggregation
checks that fail is labeled `FALSE`. Here, this concerns the records with
labels `"01"`, `"01.1"` and `"01.2"`.

We will next look at a more complicated example, but first note the following.
The `hierarchy()` function

- can handle any statistical aggregate, `sum()` is just the default;
- supports globbing and regular expressions in the child values;
- has an adjustable tolerance value for comparing observed with computed aggregates;
- has configurable behaviour for cases of missing data;
- can be applied per-group, defined by one or more grouping variables (see next example).

See the help file `?hierarchy` for specification and examples.



**A more complicated example**

Samplonia is divided in two districts, each of which is divided into several
provinces. Let us define the hierarchical code list.

```{r}
samplonia <- data.frame(
    region   = c("Agria", "Induston"
               , "Wheaton", "Greenham"
               , "Smokely", "Mudwater", "Newbay", "Crowdon")
  , parent = c(rep("Samplonia",2), rep("Agria",2), rep("Induston",4))
) 
samplonia
```
Recall the structure of the `samplonomy` dataset.
```{r}
data(samplonomy)
head(samplonomy)
```
We will check whether regions sum to their parent regions, for each period
and for each measure.

```{r}
rule <- validator(
  hierarchy(value, region, hierarchy=ref$codelist, by=list(period, measure))
)
out <- confront(samplonomy, rule, ref=list(codelist=samplonia))
summary(out)
```
We see that some aggregates add up correctly, and some don't. There is also
a warning which we should investigate.
```{r}
warnings(out)
```
If one of the groups contains a parent more than once it is not possible
to check whether child values add up to the aggregate. For this reason
the duplicated parent and all it's children are marked `FALSE`. Indeed we
find a duplicated record.
```{r}
subset(samplonomy, region  == "Induston" & 
                   period  == "2018Q2"   & 
                   measure == "export")
```
Just to see if we can remove the warning, let us remove the duplicate
and re-run the check.
```{r}
i <- !duplicated(samplonomy[c("region","period","measure")])
samplonomy2 <- samplonomy[i, ]

out <- confront(samplonomy2, rule, ref=list(codelist=samplonia))
# suppress some columns for brevity
summary(out)[1:7]
```

The `hierarchy()` function marks every record `FALSE` that is involved
in any check. This may make it hard to figure out which check it failed.
One can get more detailed information, by checking different parts
of the hierarchy in separate rules.
```{r}
rules <- validator(
   level0 = hierarchy(value, region, ref$level0, by=list(period, measure))
 , level1 = hierarchy(value, region, ref$level1, by=list(period, measure))
)
out <- confront(samplonomy2, rules
        , ref=list(level0=samplonia[1:2,], level1=samplonia[3:8,])
       )
summary(out)
```
We can now select records involved in violating the highest level
rules separately.
```{r}
violating(samplonomy2, out["level0"]) 
```
From this it appears that in 2015, the GDP for Agria is missing, and in 
2018Q3 there is no value for the export of Induston.






## General aggregates in long-form data

Checking aggregations in long-form format is more involved than for
data in wide format (as in Section \@ref(balance-equalities-and-inequalities)).


Here, we check in the `samplonomy` dataset that for each measure and each
period, the subregional data adds up to the regional data.
```{r}
rules <- validator(
  part_whole_relation(value
    , labels=region
    , whole="Samplonia"
    , part =c("Agria","Induston")
    , by=list(measure, period)
  )
)
```
The first argument of `part_whole_relation()` is the name of the variable
containing the values. Here, the column `value` from the samplonomy dataset.
The argument `labels` indicates the variable that labels parts and wholes.
Next, we define the label value that indicates a total. Here, a record with
region label `"Samplonia"` indicates a total. Under argument `part` we specify
the labels that have to add up to Samplonia, here the provinces Agria and
Induston. Note that there are more subregions in the dataset, for example the
district of Wheaton (a subregion of Agria). Since we do not specify them, these
are ignored. In the `by` argument we specify that the dataset must be split
into measure and period prior to checking the regional aggregates.


The output is one boolean value per record. For each block, defined by values
of `measure` and `period` either all values are `TRUE`, `FALSE`, or `NA`. The
latter indicates that the aggregate could not be computed because one of the
values is missing, or the computed aggregate could not be compared with the
aggregate in the data because it is missing (either the whole record may be
missing, or the value may be `NA`).
```{r}
out <- confront(samplonomy, rules)
# suppress some columns for brevity
summary(out)[1:7]
```
We can extract the truth values and then inspect the blocks with erroneous values
using standard R functionality.
```{r}
violating(samplonomy, out)
```
Recall that the rule was executed per block defined by `measure` and `period`.
Thus, the result  indicates three errors: one in the block of records defined
by `measure=="gdp"` and `period=="2015"`, also in the blocks defined by
`measure=="export"` and `period==2018Q2` or `period=="2018Q3"`. 

First, it seems that the 2015 GDP of Agria 
is missing from the data set. This turns out indeed to be the case.
```{r}
subset(samplonomy, region=="Agria" & period == "2015" & measure == "gdp")
```
Second, it can be seen that for Induston, there are two export values for
`"2018Q2"` while the export value for `"2018Q3"` is missing.


### Notes {-}

Specifying (group-wise) aggregates is fairly detailed job in the case of long
data (compare this with 
[balance checks](balance-equalities-and-inequalities)). There are a few things to keep in
mind when using this function.

- The argument `part` is optional. If not specified, every record not matching
  with `whole` will be considered a detail that is to be used to compute the total.
  In the current example this was not possible because besides Agria and Induston,
  we have other subregions.
- In the example we used literal values to specify the keys that define parts
  and wholes. It is possible to, recognize patterns, for example
  any years (4 digits) as a whole and a quarter as a part. See also the next example.
  Supported patterns include regular expressions (shown here) and globbing (see help file).
- It is important that the variables listed in `by` (if any) uniquely specify a
  single aggregate. So here, for each measure and period, the label `"Samplonia"`
  should occur at most once (if it does not occur the result will be `NA`).
- The default way to aggregate is to take the sum. You can specify other ways
  to aggregate by passing an `aggregator` argument. For example `aggregator=mean`.
- By default, the aggregate in the data is compared with the computed aggregate
  up to a tolerance of $10^{-8}$. This tolerance can be set using the `tol`
  argument.  E.g. for integer data you may want to set `tol=0`.

## Aggregates of time series in long format

We are going to check whether quarterly time series add up to the annual time
series.  This is more complicated because of two subtleties.

First there is not one fixed aggregate key, like `"Samplonia"`.  Rather, we
have a key _pattern_. Each total is defined by a period label that consists of
precisely four digits. So rather than recognizing a specific year we want to
recognize that a key represents any year. This can be done using a regular
expression of the form `"^\\d{4}$"`, where the `^` indicates 'start of string',
the `\\d{4}` indicates 'four times a digit' and `$` indicates 'end of string'.

Second, we wish to check annual totals against the sum over quarters for each region and each
measure. However, a value-combination of measure and region does not single out
a single value for `year`. For example, for the Induston export we have the following
annual data.

```{r}
subset(samplonomy, region=="Induston" & freq == "A" & measure=="export")
```
So in fact, we need to do the check _by year_ as well as by measure and region.
Fortunately, in this case it is easy to derive a variable that indicates the year
by selecting the first four characters from `period`.

```{r}
rules <- validator(part_whole_relation(value
  , labels = period
  , whole  = rx("^\\d{4}$")
  , by = list(region, substr(period,1,4), measure) 
  ))
out <- confront(samplonomy, rules)
```
We use `rx("^\\d{4}")` to tell `part_whole_relation` that this string must be
interpreted as a regular expression.  Here, we do not indicate `part` labels
explicitly: by default any record not matching `whole` will be treated as a
detail that must be used to compute the total.

```{r}
errors(out)
# suppress some columns for brevity
summary(out)[1:7]
```

We now get 9 fails and 10 missing values. We can filter out records that
have `NA` (lacking) results.
```{r}
lacking(samplonomy, out)
```
There are two blocks where the annual total could not be compared with 
the sum over quarterly series. The balance value of Crowdon is missing
for `"2014Q1"` as well as the import value of Wheaton for `"2019Q2"`.


Similarly, we can inspect the failing blocks
```{r}
violating(samplonomy, out)
```








