# Indicators

Until now we have discussed various types of data validation rules: decisions
that assign True or False values to a data frame. In some cases it is
convenient to have a continuous value that can then be used in further
assessing the data.

A practical example is the so-called [selective
editing](https://ec.europa.eu/eurostat/cros/content/selective-editing-theme_en)
approach to data cleaning. Here, each record in a data set is assigned a number
that expresses the risk a record poses for inferring a faulty conclusion.
Records are then ordered from high risk (records that both have suspicious
values and large influence on the final result) to low risk (records with
unsuspected values and little influence on the final result). Records with the
highest risk are then scrutinized by domain experts.


In `validate`, an indicator is a rule that returns an numerical value.  Just
like `validator` objects are lists of validation rules, `indicator` objects
are lists of indicator rules. Indices can be computed by confronting data with
an `indicator`, and using `add_indices`, the computed indices can be added to
the dataset. You can import, export, select, and combine `indicator` objects
in the same way as `validator` objects.



## A quick example

Here is a simple example of the workflow.

```{r}
library(validate)
ii <- indicator(
    BMI = (weight/2.2046)/(height*0.0254)^2 
  , mh  = mean(height)
  , mw  = mean(weight))
out <- confront(women, ii)
```
In the first statement we define an `indicator` object storing indicator
expressions.  Next, we confront a dataset with these indicators. The result is
an object of class `indication`. It prints as follows.
```{r}
out
```
To study the results, the object can be summarized.
```{r}
summary(out)
```
Observe that the first  indicator results in one value per record
while the second and third indicators  (`mh`, `mw`) each return a single
value. The single values are repeated when indicator values are added
to the data.
```{r}
head(add_indicators(women, out), 3)
```
The result is a data frame with indicators attached.


The columns `error` and `warning` indicate whether calculation of the
indicators was problematic. For example because the output of an indicator rule
is not numeric, or when it uses variables that do not occur in the data.  Use
`warnings(out)` or `errors(out)` to obtain the warning and error messages per
rule.


## Getting indicator values

Values can be obtained with the `values` function, or by converting to a
`data.frame`.  In this example we add a unique identifier (this is optional) to
make it easier to identify the results with data afterwards.
```{r}
women$id <- letters[1:15]
```
Compute indicators and convert to `data.frame`.
```{r}
out <- confront(women, ii,key="id")
tail( as.data.frame(out) )
```
Observe that there is no key for indicators `mh` and `mw` since these are
constructed from multiple records.


## Compare

A compare method for `indicator` objects is not yet implemented.






