# Working with validate {#sect-work}


```{r, include=FALSE}
source("chunk_opts.R")
```

```{r, echo=FALSE}
library(validate)
```

In this section we dive deeper into the the central object types used in the
package: the `validator` object type for storing lists of rules, and the
`confrontation` object type for storing the results of a validation.


## Manipulating rule sets

Validate stores rulesets into something called a `validator` object. The
`validator()` function creates such an object.
```{r}
v <- validator(speed >= 0, dist>=0, speed/dist <= 1.5)
v
```
Validator objects behave a lot like lists. For example, you can select items
to get a new `validator`. Here, we select the first and third element.
```{r}
w <- v[c(1,3)]
```
Here `w` is a new validator object holding only the first and third rule from
`v`. If not specified by the user, rules are given the default names `"V1"`,
`"V2"`, and so on. Those names can also be used for selecting rules.
```{r}
w <- v[c("V1","V3")]
```

Validator objects are reference objects. This means that if you do
```
w <- v
```
then `w` is not a copy of `v`. It is just another name for the same physical
object as `v`.  To make an actual copy, you can select everything.
```
w <- v[]
```
It is also possible to concatenate two validator objects. For example when you
read two rule sets from two files (See \@ref(sect-readfromfile)). This is done
by adding them together with `+`.
```{r}
rules1 <- validator(speed>=0)
rules2 <- validator(dist >= 0)
all_rules <- rules1 + rules2
```
An empty validator object is created with `validator()`.


If you select a single element of a validator object, an object of class 'rule'
is returned. This is the validating expression entered by the user, plus some
(optional) metadata.

```{r}
v[[3]]
```
Users never need to manipulate rule objects, but it can be convenient to
inspect them. As you see, the rules have some automatically created metadata.
In the next section we demonstrate how to retrieve and set the metadata.


## Rule metadata 

Validator objects behave a lot like lists. The only metadata in an R
list are the `names` of its elements. You can get and set names of a list
using the `names<-` function. Similarly, there are getter/setter functions
for rule metadata.


- `origin()`     : Where was a rule defined?
- `names()`      : The name per rule
- `created()`    : when were the rules created?
- `label()`      : Short description of the rule
- `description()`: Long description of the rule
- `meta()`       : Set or get generic metadata

Names can be set on the command line, just like how you would do it for
an R list.
```{r}
rules <- validator(positive_speed = speed >= 0, ratio = speed/dist <= 1.5)
rules
```
Getting and setting names works the same as for lists.
```{r}
names(rules)
names(rules)[1] <- "nonnegative_speed"
```
The functions `origin()`, `created()`, `label()`, and `description()` work in
the same way. It is also possible to add generic key-value pairs as metadata.
Getting and setting follows the usual recycling rules of R.
```{r}
# add 'foo' to the first rule:
meta(rules[1],"foo") <- 1
# Add 'bar' to all rules
meta(rules,"bar") <- "baz"
```
Metadata can be made visible by selecting a single rule:
```{r}
v[[1]]
```
Or by extracting it to a `data.frame`
```{r}
meta(v)
```

Some general information is obtained with `summary`,
```{r}
summary(v)
```
Here, some properties per _block_ of rules is given. Two rules occur in the same
block if when they share a variable. In this case, all rules occur in the same
block.

The number of rules can be requested with `length`
```{r}
length(v)
```
With `variables`, the variables occurring per rule, or over all the rules can be requested.
```{r}
variables(v)
variables(v,as="matrix")
```



## Rules in data frames

You can read and write rules and their metadata from and to data frames.  This
is convenient, for example in cases where rules are retrieved from a central
rule repository in a data base.

Exporting rules and their metadata can be done with `as.data.frame`.
```{r}
rules <- validator(speed >= 0, dist >= 0, speed/dist <= 1.5)
df <- as.data.frame(rules)
```

Reading from a data frame is done through the `.data` argument.
```{r}
rules <- validator(.data=df)
```
It is not necessary to define all possible metadata in the data frame. It is
sufficient to have three character columns, named `rule`, `name` and
`description` in any order.



## Validation rule syntax {#sect-syntax}

Conceptually, any R statement that will evaluate to a `logical` is considered a
validating statement. The validate package checks this when the user defines a
rule set, so for example calling `validator( mean(height) )` will result in a
warning since just computing `mean(x)` does not validate anything.

You will find a concise description of the syntax in the `syntax` help file.
```{r,eval=FALSE}
?syntax
```

In short, you can use

- Type checks: any function starting with `is.`
- Binary comparisons: `<, <=, ==, !=, >=, >` and `%in%`
- Unary logical operators: `!, all(), any()`
- Binary logical operators: `&, &&, |, ||` and logical implication, e.g. `if (staff > 0) staff.costs > 0`
- Pattern matching  `grepl`
- [Functional dependency](https://en.wikipedia.org/wiki/Functional_dependency): $X\to Y + Z$ is represented by `X ~ Y + Z`.

There are some extra syntax elements that help in defining complex rules.

- Inspect the whole data set using `.`, e.g. `validator( nrow(.) > 10)`.
- Reuse a variable using `:=`, e.g. `validator(m := mean(x), x < 2*m )`.
- Apply the same rule to multiple groups with `var_group`. For example 
  `validator(G:=var_group(x,y), G > 0)` is equivalent to `validator(x>0, y>0)`.

A few helper functions are available to compute groupwise values on
variables (vectors). They differ from functions like `aggregate` or `tapply`
in that their result is always of the same length as the input.
```{r}
sum_by(1:10, by = rep(c("a","b"), each=5) )
```
This is useful for rules where you want to compare individual values with
group aggregates.


|function             | computes                         |
|---------------------|----------------------------------|
| `do_by`             | generic groupwise calculation   |
| `sum_by`            | groupwise sum                    |
| `min_by`, `max_by`  | groupwise min, max               |
| `mean_by`           | groupwise mean                   |
| `median_by`         | groupwise median                 |

See also Section \@ref(sect-groupwise).

There are a number of functions that perform a particular validation task that
would be hard to express with basic syntax.  These are treated extensively
in Chapters \@ref(sect-varlevelchecks) to \@ref(sect-statisticalchecks), but
here is a quick overview.

|function             | checks                                                         |
|---------------------|----------------------------------------------------------------|
|`in_range`           | Numeric variable range                                         |
|`is_unique`          | Uniqueness of variable combinations                            |
|`all_unique`         | Equivalent to `all(is_unique())`                               |
|`is_complete`        | Completeness of records                                        |
|`all_complete`       | Equivalent to `all(is_complete())`                             |
|`exists_any`         | For each group, check if any record satisfies a rule           |
|`exists_one`         | For each group, check if exactly one record satisfies a rule   |
|`is_linear_sequence` | Linearity of numeric or date/time/period series                |
|`in_linear_sequence` | Linearity of numeric of date/time/period series                |
|`hierarchy`          | Hierarchical aggregations                                      |
|`part_whole_relation`| Generic part-whole relations                                   |
|`field_length`       | Field length                                                   |
|`number_format`      | Numeric format in text fields                                  |
|`field_format`       | Field format                                                   |
|`contains_exactly`   | Availability of records                                        |
|`contains_at_least`  | Availability of records                                        |
|`contains_at_most`   | Availability of records                                        |
|`does_not_contain`   | Correctness of key combinations                                |


## Confrontation objects

The outcome of confronting a validator object with a data set is an object of
class `confrontation`. There are several ways to extract information from a
`confrontation` object.

- `summary`: summarize output; returns a `data.frame`
- `aggregate`: aggregate validation in several ways
- `sort` : aggregate and sort in several ways
- `values`: Get the values in an array, or a list of arrays if rules have different output dimension structure
- `errors`: Retrieve error messages caught during the confrontation
- `warnings`: Retrieve warning messages caught during the confrontation.

By default aggregates are produced by rule.
```{r}
v  <- validator(height>0, weight>0,height/weight < 0.5)
cf <- confront(women, rules)
aggregate(cf) 
```
To aggregate by record, use `by='record'`
```{r}
head(aggregate(cf,by='record'))
```
Aggregated results can be automatically sorted, so records with the most violations or
rules that are violated most sort higher.
```{r}
# rules with most violations sorting first:
sort(cf)
```

Confrontation objects can be subsetted with single bracket operators (like
vectors), to obtain a sub-object pertaining only to the selected rules.
```
summary(cf[c(1,3)])
```


## Confrontation options {#sect-options}
By default, all errors and warnings are caught when validation rules are confronted with data. This can be switched off by setting the `raise` option to `"errors"` or `"all"`. The following 
example contains a specification error: `hite` should be `height` and therefore the rule errors
on the `women` data.frame because it does not contain a column `hite`. The error is caught
(not resulting in a R error) and shown in the summary,
```{r}
v <- validator(hite > 0, weight>0)
summary(confront(women, v))
```
Setting `raise` to `all` results in a R error:

```{r eval=TRUE, error=TRUE}
# this gives an error
confront(women, v, raise='all')
```


Linear equalities form an important class of validation rules. To prevent
equalities to be strictly tested, there is an option called `lin.eq.eps` (with
default value $10^{-8}$) that allows one to add some slack to these tests. The
amount of slack is intended to prevent false negatives (unnecessary failures)
caused by machine rounding. If you want to check whether a sum-rule is
satisfied to within one or two units of measurement, it is cleaner to define
two inequalities for that.








## Using reference data

For some checks it is convenient to compare the data under scrutiny with 
other data artifacts. Two common examples include:

- Data is checked against an earlier version of the same dataset.
- We wish to check the contents of a column against a code list,
  and we do not want to put the code list hard-coded into the 
  rule set.

For this, we can use the `ref` option in confront. Here is how
to compare columns from two data frames row-by-row. The user
has to make sure that the rows of the data set under scrutiny
(`women`) matches row-wise with the reference data set (`women1`).

```{r}
women1 <- women
rules <- validator(height == women_reference$height)
cf <- confront(women, rules, ref = list(women_reference = women1))
summary(cf)
```

Here is how to make a code list available.
```{r}
rules <- validator( fruit %in% codelist )
fruits <-  c("apple", "banana", "orange")
dat <- data.frame(fruit = c("apple","broccoli","orange","banana"))
cf <- confront(dat, rules, ref = list(codelist = fruits))
summary(cf)
```




