# Rules in text files {#sect:rulefiles}


```{r, include=FALSE}
source("chunk_opts.R")
library(validate)
```

This Chapter is about importing and exporting rules from and to file, both in
free-form text and in YAML. We also discuss some more advanced features like
how to have one rule file include another file.


## Reading rules from file {#sect:readfromfile}

It is a very good idea to store and maintain rule sets outside of your R
script. Validate supports two file formats: simple text files and `yaml` files.
Here we only discuss simple text files, yaml files are treated in \@ref(sect:yamlfiles).

To try this, copy the following rules into a new text file and store it in a
file called `myrules.R`, in the current working directory of your R session.

```{r, eval=FALSE}
# basic range checks
speed >= 0
dist  >= 0

# ratio check
speed / dist <= 1.5
```
Note that you are allowed to annotate the rules as you would with 
regular R code. Reading these rules can be done as follows.
```{r}
rules <- validator(.file="myrules.R")
```

## Metadata in text files: `YAML` {#sect:yamlfiles}

[YAML](https://yaml.org) is a data format that aims to be easy to learn and
human-readable. The name 'YAML' is a [recursive
acronym](https://en.wikipedia.org/wiki/Recursive_acronym) that stands for

> YAML Ain't Markup Language.

Validate can read and write rule sets from and to YAML files. For example,
paste the following code into a file called `myrules.yaml`.
```
rules:
- expr: speed >= 0
  name: 'speed'
  label: 'speed positivity'
  description: |
    speed can not be negative
  created: 2020-11-02 11:15:11
  meta:
    language: validate 0.9.3.36
    severity: error
- expr: dist >= 0
  name: 'dist'
  label: 'distance positivity'
  description: |
    distance cannot be negative.
  created: 2020-11-02 11:15:11
  meta:
    language: validate 0.9.3.36
    severity: error
- expr: speed/dist <= 1.5
  name: 'ratio'
  label: 'ratio limit'
  description: | 
    The speed to distance ratio can
    not exceed 1.5.
  created: 2020-11-02 11:15:11
  meta:
    language: validate 0.9.3.36
    severity: error

```
We can read this file using `validator(.file=)` as before.

```{r}
rules <- validator(.file="myrules.yaml")
rules
```
Observe that the labels are printed between brackets.  There are a few things
to note about these YAML files.

1. `rules:` starts a list of rules. 
2. Each new rule starts with a dash (`-`)
3. Each element of a rule is denoted `name: <content>`. The only obligated
   element is `expr`: the rule expression.
4. Spaces matter. Each element of a rule must be preceded by a newline and two spaces.
   Subelements (as in `meta`) are indented again.

A full tutorial on YAML can be found at
[W3Cschools.io](https://www.w3schools.io/file/yaml-introduction/). 


To export a rule set to yaml, use the `export_yaml()` function.
```{r}
rules1 <- rules[c(1,3)]
export_yaml(rules1, file="myrules2.yaml")
```





## Setting options

Both free-form and YAML files can optionally start with a header section where
options or file inclusions can be set. The header section is enclosed by lines
that contain three dashes (`---`) at the beginning of the line.

For example, in the following rule file we make sure that errors are not caught
but raised to run-time level, and we set the tolerance for checking  linear equalities and
inequalities to zero.

```
---
options:
  raise: errors
  lin.eq.eps: 0
  lin.ineq.eps: 0
---

turnover >= 0

staff >= 0

total.rev - profit == total.costs

```
The options you set here will be part of the `validator` object, that is
created once you read in the file. The options are valid for every
confrontation you use this validator for, unless they are overwritten during
the call to `confront()`.

The header section is interpreted as a block of YAML, so options and file
inclusions must be specified in that format.


## Including other rule files

In validate, rule files can include each other recursively. So file A can
include file B, which may include file C. This is useful for example in surveys
where the first part of the questionnaire goes to all respondents, and for the
second part, the contents of the questionnaire (and hence its variables) depend
on the respondent type. One could create files with specific rules for the
second part: one for each respondent group, and have each specific rule file
include the general rules that must hold for every respondent.
It can also be useful when different persons are responsible for different rule
sets. 

File inclusion can be set through the `include` option in the YAML header.

```
---
include:
  - petes_rules.yaml
  - nancys_rules.yaml
options:
  raise: errors
---
# start rule definitions here
```

## Exporting validator objects


There are three ways to do that. You can either write to a `yaml` file
immediately as follows
```{r, eval=FALSE}
v <- validator(height>0, weight> 0)
export_yaml(v,file="my_rules.yaml")
```
or you can get the `yaml` text string using `as_yaml`
```
cat(as_yaml(v))
```

Finally, you can convert a rule set to data frame and then export it
to a database.

```{r, eval=FALSE}
df <- as.data.frame(v)
```




















