# Rules in text files {#sect:rulefiles}


```{r, include=FALSE}
source("chunk_opts.R")
```

This Chapter is about importing and exporting rules from and to file, both in
free-form text and in YAML. We also discuss some more advanced features like
how to have one rule file include another file.

- Importing from free-form text files was discussed in \@ref{sect:readfromfile}
- Importing from simple YAML files was discussed in \@ref{sect:yamlfiles}


## Setting options

Both free-form and YAML files can optionally start with a header section where
options or file inclusions can be set. The header section is enclosed by lines
that contain three dashes (`---`) at the beginning of the line.

For example, in the following rule file we make sure that errors are not caught
but raised to run-time level, and we set the tolerance for checking  linear equalities and
inequalities to zero.

```
---
options:
  raise: errors
  lin.eq.eps: 0
  lin.ineq.eps: 0
---

turnover >= 0

staff >= 0

total.rev - profit == total.costs

```
The options you set here will be part of the `validator` object, that is
created once you read in the file. The options are valid for every
confrontation you use this validator for, unless they are overwritten during
the call to `confront()`.

The header section is interpreted as a block of YAML, so options and file
inclusions must be specified in that format.


## Including other rule files

In validate, rule files can include each other recursively. So file A can
include file B, which may inclide file C. This is useful for example in surveys
where the first part of the questionnaire goes to all respondents, and for the
second part, the contents of the questionnaire (and hence its variables) depend
on the respondent type. One could create a files with specific rules for the
second part: one for each respondent group, and have each specific rule file
include the general rules that must hold for every respondent.
It can also be useful when different persons are responsible for different rule
sets. 

File inclusion can be set through the `include` option in the YAML header.

```
---
include:
  - petes_rules.yaml
  - nancys_rules.yaml
options:
  raise: errors
---
# start rule definitions here
```

## Exporting validator objects


There are three ways to do that. You can either write to a `yaml` file
immediately as follows
```{r, eval=FALSE}
v <- validator(height>0, weight> 0)
export_yaml(v,file="my_rules.yaml")
```
or you can get the `yaml` text string using `as_yaml`
```
cat(as_yaml(v))
```

Finally, you can convert a rule set to data frame and then export it
to a database.

```{r, eval=FALSE}
df <- as.data.frame(v)
```




















