# Rules from SDMX {#sect:sdmxrules}


```{r, include=FALSE}
source("chunk_opts.R")
library(validate)
```

## SDMX and `validate`

Statistical Data and Metadata eXchange, or SDMX is a standard for storing data
and the description of its structure, meaning, and content. The standard is
developed by the [SDMX consortium](https://sdmx.org). It is used, amongst
others, in the [Official Statistics](https://en.wikipedia.org/wiki/Official_statistics) community to
exchange data in a standardized way.

A key aspect of SDMX is a standardized way to describe variables, data
structure (how is it stored), and code lists. This metadata is defined in an
_SDMX registry_ where data producers can download or query the necessary
metadata. Alternatively, metadata is distributed in a so-called Data Structure
Definition (DSD) file, which is usually in
[XML](https://en.wikipedia.org/wiki/XML) format.


For data validation, some aspects of the metadata are of interest. In
particular, code lists are interesting objects to test against.  In validate
there are two ways to use SDMX codelists. The first is by referring to a
specific code list for a specific variable in an SDMX registry. The second way
is to derive a rule set from a DSD file that can be retrieved from a registry.

Below we discuss the follwing functions.

|function            | what it does                                     |
|--------------------|--------------------------------------------------|
|`sdmx_endpoint`     | retrieve URL for SDMX endpoint                   |
|`sdmx_codelist`     | retrieve sdmx codelist                           |
|`estat_codelist`    | retrieve codelist from Eurostat SDMX registry    |
|`global_codelist`   | retrieve codelist from Global SDMX registry      |
|`validator_from_dsd`| derive validation rules from DSD in SDMX registry|


## SDMX and API locations

SDMX metadata is typically exposed through a standardized REST API.
To query an SDMX registry, one needs to supply at least the following
information:

- The registry's API entry point. This is the base URL for the online registry.
  You can specify it literally, or use one of the helper functions that 
  are aware of certan known SDMX registries.
- Agency ID: the ID of the agency that responsible for the code list
- Resource IDL the name of the SDMX resource. This is usually the name of 
  a type of statistic, like STS (short term statistics)
- version: the code list version.


Some API endpoints are stored with the package. The function `sdmx_endpoint()` 
returns endpoint URLs for several SDMX registries. Use
```{r,eval=FALSE}
sdmx_endpoint()
```
to get a list of valid endpoints. As an example, to retrieve the endpoint for the global
SDMX registry, use the following.
```{r, eval=TRUE}
sdmx_endpoint(registry="global")
```


## Code lists from SDMX registries

Code lists can be retrieved on-the-fly from one of the online
SDMX registries. In the following rule we retrieven the codelist of economic activiteis
from the [global SDMX registry](https://registry.sdmx.org/ws/public/sdmxapi).

```{r, eval=FALSE}
codelist <- sdmx_codelist(
  endpoint = sdmx_endpoint("global")
  , agency_id = "ESTAT"
  , resource_id = "CL_ACTIVITY")

head(codelist)
[1] "_T"  "_X"  "_Z"  "A"   "A_B" "A01"
```

Equivalently, and as a convenience, you could use `global_codelist()` to avoid
specifying the API endpoint explicitly. The output can be used in a rule.
```{r, eval=FALSE}
Activity %in% global_codelist(agency_id="ESTAT", resource_id="CL_ACTIVITY")
```

Since downloading codelists can take some time, any function that accesses online
SDMX registries will store the download in memory for the duration of the R session.

There is also a `estat_codelist()` function for downloading codelists from 
the Eurostat SDMX registry.




## Derive rules from DSD

The functions described in the previous subsection allow you to check
variables against a particular SDMX code list. It is also possible to
download a complete Data Structure Definition and generate all checks
implied by the DSD. 

```{r, eval=FALSE}
rules <- validator_from_dsd(endpoint = sdmx_endpoint("ESTAT")
   , agency_id = "ESTAT", resource_id = "STSALL", version="latest")

```

The `validator_from_dsd` function adds relevant metadata such as a
rule name, the origin of the rule and a short description.




